Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-06 workflow/workflow.Rmd
Updates to text and figure captions.

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any metadata available from the experiment.

@@

This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs), which become activated to differentiate and reconstitute all major cell types in the epithelium.

@@

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)). In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

@@

Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

@@

```{r scone, fig.cap="Plot generated by scone to filter low-quality cells."}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

Finally, for computational efficiency, we retain only the 1,000 most variable genes. This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]). In general, however, we recommend care in selecting an approapriate gene filtering scheme for an actual data analysis.

[FP: check. SD: I rephrased the above paragraph.]

@@

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics("zinb_schema.pdf")
```

@@

```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = "n", ylab="Expression measure")
abline(h=0)
```

@@

```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell. In the left panel, cells are color-coded by batch. In the right panel, cells are color-coded by original published clustering."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = "")
plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
```

@@

```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
W <- colData(se)[, grepl("^W", colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "bottomright", legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = "Sample")
```

@@

```{r examineCombineMany, fig.cap="Candidate clusterings found using the function RSEC from the clusterExperiment package."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

@@

```{r plotcoclust, fig.cap="Heatmap of co-clustering matrix."}
plotCoClustering(ceObj)
```

@@

```{r barplotOurs, fig.cap="Barplot of number of cells per cluster, for our workflow's clustering."}
plotBarplot(ceObj, legend = F) 
```

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)), the main difference being that several of the published clusters were merged into single clusters. This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to RSEC instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.

[FP: check. SD: Rephrased a little.]

```{r addPublishedClusters, fig.cap="Barplot of number of cells per cluster, for original published clustering."}
ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
                   clusterLabel = "publishedClusters")
plotBarplot(ceObj,whichClusters=c("combineMany","publishedClusters"), legend = F) 
```

```{r comparePublished, fig.cap="Comparison between the clustering found using our workflow and the original published clustering."}
plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
```

@@

```{r heatmapsClusters, fig.cap="Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
colData(ceObj)$publishedClusters <- as.factor(colData(ceObj)$publishedClusters)
origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$publishedClusters)]
experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors,
                   "publishedClusters" = origClusterColors)

@@

```{r mdsWce, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "bottomright", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

[SD: Check colors.]

@@

```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
pairs(lineages, type="lineages", col = pal[cl])
```

@@

```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)

@@

```{r heatmapsignificant, fig.cap="Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
heatclus <- cl[order(t, na.last = NA)]
ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
```

>>>>

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.

@@

This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons (mOSN) that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs), which become activated to differentiate and reconstitute all major cell types in the epithelium.

@@

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches off to give rise to mOSN, microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)). In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

@@

Throughout the workflow, we use the class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

@@

```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

Finally, for computational efficiency, we retain only the 1,000 most variable genes. This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]). In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.

@@

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics("zinb_schema.pdf")
```

@@

```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = "n", ylab="Expression measure")
abline(h=0)
```

@@

```{r pcanorm, fig.cap="ZINB-WaVE: PCA of normalized expression measures, where each point represents a cell. Cells are color-coded by batch (left panel) and by original published clustering (right panel)."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = "")
plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
```

@@

```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
W <- colData(se)[, grepl("^W", colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "bottomright", legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = "Sample")
```

@@

```{r examineCombineMany, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

@@

```{r plotcoclust, fig.cap="RSEC: Heatmap of co-clustering matrix."}
plotCoClustering(ceObj)
```

@@

```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering."}
plotBarplot(ceObj, legend = F) 
```

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)), the main difference being that several of the published clusters were merged here into single clusters. This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to `RSEC` instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.

```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for original published clustering."}
ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
                   clusterLabel = "publishedClusters")
plotBarplot(ceObj,whichClusters=c("combineMany","publishedClusters"), legend = F) 
```

```{r comparePublished, fig.cap="RSEC: Comparison between the clustering found using our workflow and the original published clustering."}
plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
```

@@

```{r heatmapsClusters, fig.cap="RSEC: Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
colData(ceObj)$publishedClusters <- as.factor(colData(ceObj)$publishedClusters)
origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$publishedClusters)]
experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors,
                   "publishedClusters" = origClusterColors)

@@

```{r mdsWce, fig.cap="RSEC: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "bottomright", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

@@

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred global lineage structure."}
pairs(lineages, type="lineages", col = pal[cl])
```

@@

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)

@@

```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
heatclus <- cl[order(t, na.last = NA)]
ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
```
